---
title: "scRNA-seq Part 2b: Clustering and Marker Discovery"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
params:
  config.args: "./config.yaml"
---


```{r setup}
knitr::opts_chunk$set(
  warning=FALSE,error=FALSE, message=TRUE, echo=FALSE
)

source("functions_for_report.R") # This is the file that holds the R functions used in the RMD file
```



```{r libraries}
if (!suppressMessages(require(Seurat))) stop("Seurat is required")
if (!suppressMessages(require(yaml))) stop("yaml is required")
if (!suppressMessages(require(cluster))) stop("cluster is required") # For silhouette score calculation
if (!suppressMessages(require(future))) stop("future is required")
if (!suppressMessages(require(future.apply))) stop("future.apply is required")
if (!suppressMessages(require(cowplot))) stop("cowplot is required")
if (!suppressMessages(require(parallelly))) stop("parallelly is required")
if (!suppressMessages(require(ggplot2))) stop("ggplot2 is required")
if (!suppressMessages(require(dplyr))) stop("dplyr is required")
if (!suppressMessages(require(kableExtra))) stop("kableExtra is required")


# the presto lib isn't required, but FindAllMarkers runs MUCH faster with it. 
# For example, for 80k cells and 10 clusters, without presto FindAllMarkers 
# took more than 5 hours, but with presto loaded it took less than 
# 5 minutes!!
#
# The latest ragg, which is a dependency for devtools, which is itself required
# for presto install, doesn't work with R/4.4.0 (as of 13-Nov-2025), so install 
# an older version of ragg, then install devtools, then install presto
# 
# install.packages(
#   "https://cran.r-project.org/src/contrib/Archive/ragg/ragg_1.4.0.tar.gz",
#   repos = NULL,
#   type = "source"
# )
# install.packages(devtools)
# devtools::install_github("immunogenomics/presto")
if (!requireNamespace("presto", quietly = TRUE)) {
  message("Warning: presto not installed — FindAllMarkers will take a long time to run.")
}


```

```{r prepare_inline_variables}
# get the datetime and set it to variable
todays.date <- Sys.Date()
formatted.date <- format(todays.date, format="%B %d %Y")

config.args <- read_yaml(params$config.args)

```

```{r user_vals}
cfg <- config.args
cfg.analysis <- cfg$analysis

p1  <- cfg.analysis$part1
p2 <- cfg.analysis$part2

r.obj.loc <- cfg$data$`rds-file-path`
organism <- cfg$project$organism

################################################################################
# QC cutoffs
################################################################################

mito.ceiling = p2$mito_ceiling
RNA.count.floor = p2$RNA_count_floor
RNA.count.ceiling = p2$RNA_count_ceiling
feature.count.floor = p2$feature_count_floor
feature.count.ceiling = p2$feature_count_ceiling
rbc.ceiling = p2$rbc_ceiling
ribo.floor = p2$ribo_floor

node.type <- cfg.analysis$node_type
run.doubletFinder <- as.logical(p2$run_doubletFinder)

if (run.doubletFinder){
  if (!suppressMessages(require(DoubletFinder))) stop("DoubletFinder is required")
}

################################################################################
# I/O paths
################################################################################
part1.suffix <- p1$part1_suffix
part2.suffix <- p2$part2_suffix

unfiltered.seurat.obj.path    <- suffix_path(r.obj.loc, p1$part1_rds_save_filename,  part1.suffix)
merged.seurat.obj.path <- suffix_path(r.obj.loc, p2$merged_seurat_obj_file, part2.suffix)
integrated.seurat.obj.path <- suffix_path(r.obj.loc, p2$integrated_seurat_obj_file, part2.suffix)
QC.tables.path         <- suffix_path(r.obj.loc, p1$part1_report_tables_filename, part1.suffix)
ggplot.directory       <- file.path(r.obj.loc, p1$ggplot_dir)
marker.gene.dir <- suffix_path(r.obj.loc, p2$marker_gene_dir, part2.suffix)

  if (!file.exists(ggplot.directory)) {
 dir.create(ggplot.directory, showWarnings = FALSE, recursive = TRUE)
}

if (!file.exists(marker.gene.dir)) {
 dir.create(marker.gene.dir, showWarnings = FALSE, recursive = TRUE)
}


```

```{r get_num_cores_available}
if (node.type == "compute"){
  # If using a compute node, then put this in the submission script:
  # export MC_CORES=${SLURM_NTASKS}
  num.cores.available <- as.numeric(Sys.getenv("SLURM_NTASKS"))
} else {
  library(parallel)
  num.cores.available <- detectCores()
}

# Parallelism is intentionally left SEQUENTIAL during PDF rendering: forked/
# multisession futures are the usual cause of the "never-ending render" noted in
# the README. FindAllMarkers speed comes from presto, not from a future plan.
# To parallelize interactively, replace the line below with, e.g.:
#   future::plan("multicore", workers = num.cores.available)
message("Cores detected: ", num.cores.available, " (render uses sequential plan)")
future::plan("sequential")
```

## Inputs and staleness check

Part 2b consumes the integrated object produced by part 2a. It re-clusters at the
resolution recorded in the config (`clustering_resolution_value`), which you can
override by hand, then runs marker discovery and plots.

```{r guard_and_load}
# 1) The integrated object must exist.
if (!file.exists(integrated.seurat.obj.path)) {
  stop("Integrated object not found at ", integrated.seurat.obj.path,
       "\nRun part 2a first (or check part1/part2 suffixes in the config).")
}

seurat5.obj.integrated <- readRDS(integrated.seurat.obj.path)

# 2) The object must have been built under the CURRENT upstream settings.
#    If QC cutoffs / doublet flag / organism / suffixes changed since 2a ran,
#    this object is stale and 2a must be re-run.
current.params <- upstream_params(
  p2 = p2, run.doubletFinder = run.doubletFinder,
  organism = organism, part1.suffix = part1.suffix, part2.suffix = part2.suffix
)
stop_if_stale(seurat5.obj.integrated, current.params)

# 3) The clustering graph must be present.
if (!("int_snn" %in% names(seurat5.obj.integrated@graphs))) {
  stop("The 'int_snn' graph is missing from the integrated object. ",
       "Re-run part 2a, which builds it before saving.")
}
```

## Clustering

```{r run_clustering}
# Resolution comes from the config value (proposed by 2a, or your override).
if (is.null(p2$clustering_resolution_value)) {
  stop("clustering_resolution_value is not set in the config. ",
       "Run part 2a to propose one, or set it by hand.")
}
best_res <- as.numeric(p2$clustering_resolution_value)
cat("Clustering at resolution:", best_res, "\n")

seurat5.obj.integrated <- FindClusters(seurat5.obj.integrated,
                                       graph.name = "int_snn",
                                       resolution = best_res,
                                       verbose = FALSE)

cluster.res.prefix <- "int_snn_res."
clustering.resolution <- paste0(cluster.res.prefix, best_res)
num.clusters <- length(unique(Idents(seurat5.obj.integrated)))
cat("Found", num.clusters, "clusters.\n")

################################################################################
# Set the cluster numbers to be numeric so that they get printed out in sorted
# order when printed in tables, figures, etc
################################################################################
current.idents <- Idents(seurat5.obj.integrated)
correct.levels <- as.character(sort(as.integer(levels(current.idents))))
Idents(seurat5.obj.integrated) <- factor(current.idents, levels = correct.levels)

new.clusters <- factor(
  as.character(seurat5.obj.integrated[[clustering.resolution]][, 1]),
  levels = correct.levels
)
seurat5.obj.integrated <- AddMetaData(seurat5.obj.integrated,
                                       metadata = new.clusters,
                                       col.name = clustering.resolution)

saveRDS(seurat5.obj.integrated, integrated.seurat.obj.path, compress = FALSE)
cat(paste0("Saved clustered object (resolution ", clustering.resolution,
           ") to ", integrated.seurat.obj.path, "\n"))

################################################################################
# Write derived values back to config (authoritative label + cluster count).
# clustering_resolution_value is left as-is so the knob the user set is preserved.
################################################################################
cfg$analysis$part2$clustering_resolution <- clustering.resolution
cfg$analysis$part2$num_clusters <- num.clusters
yaml::write_yaml(cfg, params$config.args)
```

## Table of cells from each sample per cluster

```{r table_cells_cluster_sample}

cluster.sample.table <- table(
  Cluster = Idents(seurat5.obj.integrated),
  Sample  = seurat5.obj.integrated$sample_id
)
cluster.sample.df <- as.data.frame.matrix(cluster.sample.table)

# Add row totals
cluster.sample.df$Total <- rowSums(cluster.sample.df)

# Add column totals
totals.row <- c(colSums(cluster.sample.df))
cluster.sample.df <- rbind(cluster.sample.df, Total = totals.row)

# Add cluster column
cluster.sample.df <- cbind(Cluster = c(rownames(cluster.sample.df)[-nrow(cluster.sample.df)], "Total"),
                            cluster.sample.df)
rownames(cluster.sample.df) <- NULL
cluster.sample.df

write.csv(cluster.sample.df, file.path(r.obj.loc, "cells_perCluster_perSample.csv"), quote=FALSE)
```

## Umap of the Seurat object after integration, colored by sample

```{r integrated_umap_by_sample}
integrated.dimplot.by.sample <- DimPlot(seurat5.obj.integrated,
                                        pt.size = 0.001,
                                        reduction = "umap.postint",
                                        shuffle = TRUE,
                                        group.by = "sample_id")


# If you want to see this plot faceted by sample, i.e. one plot per sample, 
# uncomment this code:

# integrated.dimplot.by.sample <- DimPlot(seurat5.obj.integrated, 
#                                         pt.size = 0.001, 
#                                         reduction = "umap.postint", 
#                                         group.by = "sample_id", 
#                                         split.by = "sample_id", 
#                                         ncol = 2) + NoLegend()

save_png_plot(integrated.dimplot.by.sample, "integrated_umap_by_sample.png", ggplot.directory)
integrated.dimplot.by.sample

```
## Umap of the Seurat object after integration, colored by clusters 

```{r integrated_umap_by_cluster}
integrated.dimplot.by.cluster <- DimPlot(seurat5.obj.integrated, 
                              reduction = "umap.postint", 
                              group.by = clustering.resolution,
                              label = TRUE)

save_png_plot(integrated.dimplot.by.cluster, "integrated_umap_by_cluster.png", ggplot.directory)
integrated.dimplot.by.cluster
```

## UMAP of PCA

```{r pca_plots}
pca_post <- DimPlot(seurat5.obj.integrated, reduction = "pca", group.by = "sample_id", label = TRUE)


pca_post
save_png_plot(pca_post, "pca_integrated.png", ggplot.directory)
```

# Find Markers

FindAllMarkers for clusters. 

```{r find_markers_sct}
DefaultAssay(seurat5.obj.integrated) <- "SCT"
Idents(seurat5.obj.integrated) <- clustering.resolution
# perform differential expression

seurat5.obj.integrated <- PrepSCTFindMarkers(seurat5.obj.integrated)

# Runs MUCH faster with presto package installed
cluster.markers.sct <- FindAllMarkers(seurat5.obj.integrated, only.pos = TRUE, verbose = TRUE)

saveRDS(cluster.markers.sct, file=file.path(marker.gene.dir, "cluster_markers_sct.RDS"))
# cluster.markers.sct <- readRDS(file.path(marker.gene.dir, "cluster_markers_sct.RDS"))

################################################################################
# Save markers
################################################################################
# Iterate over the clusters that ACTUALLY appear in the marker table, in numeric
# order. The previous version derived a count from max(cluster id) and looped
# seq(0, n-1), which silently dropped the highest-numbered cluster (and any
# cluster with no positive markers). Note: with only.pos = TRUE, a cluster with
# no up-regulated markers will legitimately have no file.
clusters.present <- sort(unique(as.integer(as.character(cluster.markers.sct$cluster))))
for (i in clusters.present){
  table.filename <- file.path(marker.gene.dir, paste0("cluster", i, "_marker_genes_", clustering.resolution, ".csv"))
  these.markers <- cluster.markers.sct[cluster.markers.sct$cluster == i,]
  write.csv(x = these.markers, file = table.filename, quote = FALSE)
}

write.csv(cluster.markers.sct, 
          file.path(marker.gene.dir, paste0("cluster_markers_", clustering.resolution, ".csv")))

saveRDS(seurat5.obj.integrated, integrated.seurat.obj.path, compress=FALSE)

print("Saved final object, with prepSCTFindMarkers performed.")
```

```{r find_markers_RNA, eval = FALSE}
# Just in case someone wants to look for markers in RNA assay
DefaultAssay(seurat5.obj.integrated) <- "RNA"
Idents(seurat5.obj.integrated) <- clustering.resolution

seurat5.obj.integrated <- NormalizeData(seurat5.obj.integrated, 
                                        normalization.method = "LogNormalize", 
                                        scale.factor = 1e4, verbose = FALSE)

seurat5.obj.integrated <- FindVariableFeatures(seurat5.obj.integrated, 
                                               selection.method = "vst", 
                                               nfeatures = 2000, verbose = FALSE)

cluster.markers.RNA <- FindAllMarkers(seurat5.obj.integrated, assay = "RNA", test.use = "wilcox", verbose = TRUE)
saveRDS(cluster.markers.RNA, file=file.path(marker.gene.dir, "cluster_markers_RNA.RDS"))
```

## Plot markers

```{r gene_feature_plots}
markers <- cluster.markers.sct # or cluster.markers.RNA
num.clusters.list <- sort(unique(markers$cluster))   # sorted, not hash order

# Which marker plots to draw. Config-driven; default feat + dot.
marker.plot.types <- p2$marker_plot_types %||% c("feat", "dot")

invisible(lapply(num.clusters.list,
                 feat_plots_top_genes,
                 seurat.obj      = seurat5.obj.integrated,
                 marker.genes.df = markers,
                 ggplot.dir      = ggplot.directory,
                 filename.suffix = part2.suffix,
                 plots           = marker.plot.types,
                 assay.used      = "SCT",
                 num_cols        = 3,
                 to_raster       = FALSE))

```


## Session Information
```{r session_info, eval=TRUE}
sessionInfo()
```



```{r filter_clusters, eval=FALSE}
################################################################################
# In some cases, there are 'clusters' identified where there are only a few 
# cells in them and there are no markers discovered. Here is how to remove those 
# clusters. Note that this assumes that the clusters to be removed are the 
# highest-numbered clusters. See the commented-out code at the bottom if that's 
# not the case.
#
# Also, make sure to rerun the above chunk "integrated_umap_by_cluster" to 
# make a fresh umap of the clustering
################################################################################

# Identify clusters with too few cells
min.cluster.size <- 5
cluster.sizes <- table(Idents(seurat5.obj.integrated))
small.clusters <- names(cluster.sizes[cluster.sizes <= min.cluster.size])
cat("Removing clusters with <=", min.cluster.size, "cells:", paste(small.clusters, collapse = ", "), "\n")
cat("Total cells to remove:", sum(cluster.sizes[small.clusters]), "\n")

# Remove them
seurat5.obj.integrated <- subset(seurat5.obj.integrated, 
                                  idents = small.clusters, 
                                  invert = TRUE)
cat("Cells remaining:", ncol(seurat5.obj.integrated), "\n")

# After removing them, re-run the cluster factor reordering since the levels 
# will still reference the removed clusters:

current.idents <- Idents(seurat5.obj.integrated)
correct.levels <- as.character(sort(as.integer(levels(droplevels(current.idents)))))
Idents(seurat5.obj.integrated) <- factor(current.idents, levels = correct.levels)

# Update the config
num.clusters <- length(levels(Idents(seurat5.obj.integrated)))
cfg$analysis$part2$num_clusters <- num.clusters

yaml::write_yaml(cfg, params$config.args)

# And the object
saveRDS(seurat5.obj.integrated, integrated.seurat.obj.path, compress=FALSE)


################################################################################
# Note that the code above relies on the assumption that the clusters being 
# removed are the highest-numbered clusters, but if that's not the case
# then you'll need to re-map the cluster IDs and adjust the marker genes table:
################################################################################

# # Remap cluster IDs to be contiguous after removal
# old.levels <- levels(Idents(seurat5.obj.integrated))
# new.levels <- as.character(seq(0, length(old.levels) - 1))
# new.idents <- factor(
#   new.levels[as.integer(Idents(seurat5.obj.integrated))],
#   levels = new.levels
# )
# Idents(seurat5.obj.integrated) <- new.idents
# seurat5.obj.integrated <- AddMetaData(seurat5.obj.integrated,
#                                        metadata = new.idents,
#                                        col.name = clustering.resolution)
# 
# # Update markers table to match new cluster IDs
# old.to.new <- setNames(new.levels, old.levels)
# markers$cluster <- old.to.new[as.character(markers$cluster)]

```

