---
title: "scRNA-seq QC Report"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    includes:
      in_header: common-pdf.tex
params:
  config.args: "./config.yaml"
---



```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning=FALSE,error=FALSE, message=FALSE
)

# source helpers that sit beside this Rmd
source("functions_for_report.R")

```



```{r import_libraries, eval=TRUE}
if (!suppressMessages(require(Seurat))) stop("Seurat is required")
if (!suppressMessages(require(ggplot2))) stop("ggplot2 is required")
if (!suppressMessages(require(yaml))) stop("yaml is required")
if (!suppressMessages(require(knitr))) stop("knitr is required")
if (!suppressMessages(require(kableExtra))) stop("kableExtra is required")
if (!suppressMessages(require(dplyr))) stop("dplyr is required")
if (!suppressMessages(require(scales))) stop("scales is required")

```


```{r user_vars, eval=TRUE}

# get the datetime and set it to variable for inline printing
todays.date <- Sys.Date()
formatted.date <- format(todays.date, format="%B %d %Y")

# Import the config file
config.args <- read_yaml(params$config.args)

run_metrics_tables <- as.logical(config.args$analysis$part1$generate_metrics_tables)

################################################################################
# set project variables
################################################################################
analyst.name <- config.args$project$`analyst-name`
formatted.samples <- config.args$project$samples # Sanitized below
experiment.name = config.args$project$`project-name`
project.description <- config.args$project$`project-description`
dataset.path <- config.args$data$parent_directory_path
organism = config.args$project$organism
run.soupx <- as.logical(config.args$analysis$part1$run_SoupX)
if (run.soupx && !suppressMessages(require(SoupX)))
  stop("SoupX is required when run_SoupX is TRUE")

################################################################################
# Output paths. Part 1 writes into part1[_<part1_suffix>]/; later stages nest
# inside it. See stage_dirs() in functions_for_report.R.
################################################################################
dirs  <- stage_dirs(config.args, create = "part1")
p1    <- config.args$analysis$part1

unfiltered.seurat.obj.path <- file.path(dirs$part1, p1$part1_rds_save_filename)
QC.tables.path             <- file.path(dirs$part1, p1$part1_report_tables_filename)
ggplot.directory           <- file.path(dirs$part1, p1$ggplot_dir)
dir.create(ggplot.directory, showWarnings = FALSE, recursive = TRUE)


################################################################################
# Set gene subset markers (mito genes, ribo genes, etc)
################################################################################

organism <- tolower(organism)
if (!organism %in% c("human", "mouse"))
  stop("project:organism must be 'human' or 'mouse', got: '", organism, "'. ",
       "Other species need their own mito/ribo/hb patterns and Y-gene list.")

if (organism == "human") {
  mt.marker   <- "^MT-"
  hb.marker   <- "^HB[^(P)]"
  ribo.marker <- "^RP[SL]"
  y.genes     <- readLines("y_genes_hg38_symbols.txt")
} else {
  mt.marker   <- "^mt-"
  hb.marker   <- "^Hb[^(p)]"
  ribo.marker <- "^Rp[sl]"
  y.genes     <- readLines("y_genes_mm10_symbols.txt")
}


################################################################################
# Format the sample names
################################################################################

ids <- sanitize_sample_names(formatted.samples)
# seurat.sample.names <- gsub("-", "_", formatted.samples)
```


## Report and variable details

This report was generated on `r formatted.date`

* organism: `r organism`

* starting with raw data at \nolinkurl{`r dataset.path`}
* saving rds object to \nolinkurl{`r unfiltered.seurat.obj.path`}
* saving file for report at \nolinkurl{`r QC.tables.path`}
* SoupX ambient RNA correction: `r run.soupx`

## Project Description:

`r project.description`


## Sample Names

```{r sample_name_list, results='asis'}
# One per line. `formatted.samples` is what the user typed; `ids` is the
# sanitized form used for directory lookup and cell-name prefixes. They are
# usually identical - annotate only when sanitizing actually changed something,
# so a mismatch between the config and the folder on disk is visible here
# rather than surfacing later as a "could not locate 10x matrix" error.
changed <- formatted.samples != ids
cat(paste0("* ", formatted.samples,
           ifelse(changed, paste0(" (read as `", ids, "`)"), "")),
    sep = "\n")
cat("\n")
```


# Part 1: Loading data from CellRanger into R

## Sequencing QC table

For more insight on what the different metrics presented here mean, see this article on the 10x website: https://www.10xgenomics.com/analysis-guides/quality-assessment-using-the-cell-ranger-web-summary

<br/>
<br/>

```{r prep_qc_table, eval=run_metrics_tables, message=FALSE, warning=FALSE}
# --- Import the metrics summaries (quiet) ---
qc.table <- make_qc_table(ids, dataset.path)
render_formatted_table(qc.table)
```

## Load the Cell Ranger Matrix Data and create the base Seurat object.

<!-- Import the 10x data (either matrices or h5) and create a merged seurat object with counts of each sample as layers. -->

```{r soupx_explainer, eval=run.soupx, results='asis'}
cat(c(
"### Ambient RNA correction (SoupX)",
"",
"One plot per sample, in the order listed above, each titled with its sample",
"name. Each is an independent estimate of that sample's *contamination",
"fraction* (rho): the proportion of counts in a droplet that came from",
"free-floating RNA in the cell suspension rather than from the cell captured in",
"that droplet.",
"",
"**Reading them:** a single sharp peak means the candidate estimators agreed and",
"the estimate is trustworthy. A broad, flat, or multi-peaked curve means they",
"did not, and the correction for that sample deserves more scepticism. A point",
"estimate sitting at the bottom edge of its own interval means contamination is",
"at or below what the method can resolve, and the correction is close to a",
"no-op. Estimates above roughly 10-20% are worth raising with whoever prepared",
"the sample rather than simply correcting away.",
"",
"Corrected counts replace the raw counts from here on; the QC distributions",
"below are computed after this adjustment.",
""), sep = "\n")
```

```{r make_seurat_objs}
# To avoid any future.globals.maxSize errors:
options(future.globals.maxSize = Inf)
future::plan("sequential")

# make a Seurat object per sample
seurat.list <- lapply(ids, function(s) {
  m <- read_10x_any(s, dataset.path, run.soupx)
  s_fixed <- gsub("-", "_", s)
  # read_10x_any() prefers the RAW 10x matrix, which includes empty droplets.
  # min.features here only strips those empties so the QC plots are renderable;
  # it is deliberately well below the analytical feature floor
  # (feature_count_floor, applied in part 2) so the part-1 "raw" distributions
  # stay honest rather than being pre-cut at the downstream threshold.
  # Cell-name uniqueness is handled by add.cell.ids at the merge step below,
  # so no manual barcode renaming is needed here.
  o <- CreateSeuratObject(counts = m, project = experiment.name, min.features = 100)
  o$sample_id <- s_fixed
  o
})
names(seurat.list) <- gsub("-", "_", ids)
```

```{r set_percent_feats}

for (i in seq_along(seurat.list)) {
  seurat.list[[i]][["percent.mt"]] <- PercentageFeatureSet(seurat.list[[i]], pattern = mt.marker)
  # Calculate percentage of counts from hemoglobin genes (RBC markers)
  seurat.list[[i]][["percent.hb"]] <- PercentageFeatureSet(seurat.list[[i]], pattern = hb.marker)

  # Filter out the y.genes that aren't in the object
  y.genes.present <- intersect(y.genes, rownames(seurat.list[[i]]))
  if (length(y.genes.present) < length(y.genes)) {
    message(length(y.genes) - length(y.genes.present), " Y-chr genes not found in sample ",
            names(seurat.list)[i], " and will be excluded from percent.y calculation")
  }

  seurat.list[[i]][["percent.y"]] <- PercentageFeatureSet(seurat.list[[i]], features = y.genes.present)
  seurat.list[[i]][["percent.ribo"]] <- PercentageFeatureSet(seurat.list[[i]], pattern = ribo.marker)

}

```

```{r gather_qc_data_for_tables}
# Alphabetical sorting (default)
nCount_data       <- make_metric_table_from_list(seurat.list, "nCount_RNA")
nFeature_data     <- make_metric_table_from_list(seurat.list, "nFeature_RNA")
percent_mt_data   <- make_metric_table_from_list(seurat.list, "percent.mt")
percent_hb_data   <- make_metric_table_from_list(seurat.list, "percent.hb")
percent_y_data    <- make_metric_table_from_list(seurat.list, "percent.y")
percent_ribo_data <- make_metric_table_from_list(seurat.list, "percent.ribo")

# Or with a custom sample order
# sample_order <- c("sample_1", "sample_2",
#                   "sample_3", "sample_4")
#
# nCount_data <- make_metric_table_from_list(seurat.list, "nCount_RNA", sample_levels = sample_order)



to.save.for.report <- c("nCount_data",
                        "nFeature_data",
                        "percent_mt_data",
                        "percent_hb_data",
                        "percent_y_data",
                        "percent_ribo_data")

save(list = to.save.for.report, file=QC.tables.path)
```

# Part 2: QC plots

Raw distributions across all samples. Violins are width-normalized so a few
extreme cells don't flatten the rest; count metrics are on a log scale. Use
these to pick the cutoffs written into `part2:` in the config.

## Library size and complexity

```{r qc_survey_counts}
qc.counts.raw <- qc_long(list(nCount_RNA   = nCount_data,
                              nFeature_RNA = nFeature_data))

p.qc.counts <- qc_facet_plot(qc.counts.raw, log.y = TRUE)
p.qc.counts

save_png_plot(p.qc.counts, plot_path(ggplot.directory, "qc_counts_raw_facet.png"),
              width = 7, height = 3.6)
```

## Contamination and composition metrics

```{r qc_survey_percents}
qc.pct.raw <- qc_long(list(percent.mt   = percent_mt_data,
                           percent.hb   = percent_hb_data,
                           percent.ribo = percent_ribo_data,
                           percent.y    = percent_y_data))

p.qc.pct <- qc_facet_plot(qc.pct.raw, facet.nrow = 2)
p.qc.pct
save_png_plot(p.qc.pct, plot_path(ggplot.directory, "qc_percent_raw_facet.png"),
              width = 7, height = 5.5)
```


```{r merge_list_into_agg_obj}
# merge => v5 stores each sample as its own layer(s) inside RNA
# seurat.obj.v5 <- Reduce(function(x, y) merge(x, y, add.cell.ids = c(unique(x$sample_id), unique(y$sample_id))), seurat.list)

seurat.obj.v5 <- merge(
  x            = seurat.list[[1]],
  y            = seurat.list[-1],
  add.cell.ids = ids,
  project      = experiment.name
)

seurat.obj.v5 <- JoinLayers(seurat.obj.v5)

# The layer names are now defaults, e.g. counts.1, counts.2, etc. Rename the layers
# with sample names

# 1) make sure your per-cell sample label is exactly what you want in the layer names
#    (factor labels become the suffix after 'counts.')
seurat.obj.v5$sample_id <- factor(
  seurat.obj.v5$sample_id,
  levels = unique(seurat.obj.v5$sample_id),                  # or your custom order
  labels = make.names(unique(seurat.obj.v5$sample_id))       # sanitize: replace -/spaces with _
)

# 2) split the assay into layers by that factor
seurat.obj.v5[["RNA"]] <- split(seurat.obj.v5[["RNA"]], f = seurat.obj.v5$sample_id)

```


```{r save_obj}
saveRDS(seurat.obj.v5, unfiltered.seurat.obj.path)
```



## Session Information
```{r sessioninfo, eval=TRUE}
sessionInfo()
```


