---
title: "scRNA-seq Part 2a: Filtering, Integration, and Resolution Recommendation"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
params:
  config.args: "./config.yaml"
---

```{r setup}
knitr::opts_chunk$set(
  warning=FALSE,error=FALSE, message=TRUE, echo=FALSE
)

source("functions_for_report.R") # This is the file that holds the R functions used in the RMD file
```



```{r libraries}
if (!suppressMessages(require(Seurat))) stop("Seurat is required")
if (!suppressMessages(require(yaml))) stop("yaml is required")
if (!suppressMessages(require(cluster))) stop("cluster is required") # For silhouette score calculation
if (!suppressMessages(require(future))) stop("future is required")
if (!suppressMessages(require(future.apply))) stop("future.apply is required")
if (!suppressMessages(require(cowplot))) stop("cowplot is required")
if (!suppressMessages(require(parallelly))) stop("parallelly is required")
if (!suppressMessages(require(ggplot2))) stop("ggplot2 is required")
if (!suppressMessages(require(dplyr))) stop("dplyr is required")
if (!suppressMessages(require(kableExtra))) stop("kableExtra is required")


# the presto lib isn't required, but FindAllMarkers runs MUCH faster with it. 
# For example, for 80k cells and 10 clusters, without presto FindAllMarkers 
# took more than 5 hours, but with presto loaded it took less than 
# 5 minutes!!
#
# The latest ragg, which is a dependency for devtools, which is itself required
# for presto install, doesn't work with R/4.4.0 (as of 13-Nov-2025), so install 
# an older version of ragg, then install devtools, then install presto
# 
# install.packages(
#   "https://cran.r-project.org/src/contrib/Archive/ragg/ragg_1.4.0.tar.gz",
#   repos = NULL,
#   type = "source"
# )
# install.packages(devtools)
# devtools::install_github("immunogenomics/presto")
if (!requireNamespace("presto", quietly = TRUE)) {
  message("Warning: presto not installed — FindAllMarkers will take a long time to run.")
}


```

```{r prepare_inline_variables}
# get the datetime and set it to variable
todays.date <- Sys.Date()
formatted.date <- format(todays.date, format="%B %d %Y")

config.args <- read_yaml(params$config.args)

```

```{r user_vals}
cfg <- config.args
cfg.analysis <- cfg$analysis

p1  <- cfg.analysis$part1
p2 <- cfg.analysis$part2

r.obj.loc <- cfg$data$`rds-file-path`
organism <- cfg$project$organism

################################################################################
# QC cutoffs
################################################################################

mito.ceiling = p2$mito_ceiling
RNA.count.floor = p2$RNA_count_floor
RNA.count.ceiling = p2$RNA_count_ceiling
feature.count.floor = p2$feature_count_floor
feature.count.ceiling = p2$feature_count_ceiling
rbc.ceiling = p2$rbc_ceiling
ribo.floor = p2$ribo_floor

node.type <- cfg.analysis$node_type
run.doubletFinder <- as.logical(p2$run_doubletFinder)

if (run.doubletFinder){ 
  if (!suppressMessages(require(DoubletFinder))) stop("DoubletFinder is required")
}

################################################################################
# I/O paths
################################################################################
part1.suffix <- p1$part1_suffix
part2.suffix <- p2$part2_suffix

unfiltered.seurat.obj.path    <- suffix_path(r.obj.loc, p1$part1_rds_save_filename,  part1.suffix)
merged.seurat.obj.path <- suffix_path(r.obj.loc, p2$merged_seurat_obj_file, part2.suffix)
integrated.seurat.obj.path <- suffix_path(r.obj.loc, p2$integrated_seurat_obj_file, part2.suffix)
QC.tables.path         <- suffix_path(r.obj.loc, p1$part1_report_tables_filename, part1.suffix)
part2.tables.path <- suffix_path(r.obj.loc, p2$part2_report_tables_filename, part2.suffix)

ggplot.directory       <- file.path(r.obj.loc, p1$ggplot_dir)
marker.gene.dir <- suffix_path(r.obj.loc, p2$marker_gene_dir, part2.suffix)

  if (!file.exists(ggplot.directory)) {
 dir.create(ggplot.directory, showWarnings = FALSE, recursive = TRUE)
}

if (!file.exists(marker.gene.dir)) {
 dir.create(marker.gene.dir, showWarnings = FALSE, recursive = TRUE)
}


```


## Report and variable details

This report was generated on `r formatted.date`

* node.type: `r node.type`

* starting with rds object at  \nolinkurl{`r unfiltered.seurat.obj.path`}
* saving merged object to  \nolinkurl{`r merged.seurat.obj.path`}
* saving final clustered, integrated object to  \nolinkurl{`r integrated.seurat.obj.path`}

* mito ceiling: `r mito.ceiling`
* RNA count floor: `r RNA.count.floor`
* RNA count ceiling: `r RNA.count.ceiling`
* Feature count floor: `r feature.count.floor`
* Feature count ceiling: `r feature.count.ceiling`
* RBC ceiling: `r rbc.ceiling`
* Ribosomal floor: `r ribo.floor`

* Running doubletFinder: `r run.doubletFinder`


```{r get_num_cores_available}
if (node.type == "compute"){
  # If using a compute node, then put this in the submission script:
  # export MC_CORES=${SLURM_NTASKS}
  num.cores.available <- as.numeric(Sys.getenv("SLURM_NTASKS"))
} else {
  library(parallel)
  num.cores.available <- detectCores()
}

# Parallelism is intentionally left SEQUENTIAL during PDF rendering: forked/
# multisession futures are the usual cause of the "never-ending render" noted in
# the README. FindAllMarkers speed comes from presto, not from a future plan.
# To parallelize interactively, replace the line below with, e.g.:
#   future::plan("multicore", workers = num.cores.available)
message("Cores detected: ", num.cores.available, " (render uses sequential plan)")
future::plan("sequential")
```


```{r load_data}
seurat5.obj <- readRDS(file=unfiltered.seurat.obj.path)

load(QC.tables.path)

set.seed(12345)
```


<br/>


## Cell filtering stats

We use the cutoffs above to filter out cells.

```{r apply_filtering }
before.table <- table(seurat5.obj$sample_id)
md.pre <- seurat5.obj[[]]   

seurat5.obj <- subset(seurat5.obj, percent.mt <= mito.ceiling &
                                 nCount_RNA >= RNA.count.floor &
                                 nCount_RNA <= RNA.count.ceiling &
                                 nFeature_RNA >=feature.count.floor &
                                 nFeature_RNA <= feature.count.ceiling &
                                 percent.hb <= rbc.ceiling &
                                 percent.ribo >= ribo.floor)

after.table <- table(seurat5.obj$sample_id)

combo.table <- rbind(before.table,
                     after.table)

row.names(combo.table) <- c("Number of cells before filtering", "Number of cells after filtering")

combo.table
write.csv(combo.table, file.path(r.obj.loc, "filtering_stats_table.csv"), quote=FALSE)

```

```{r cells_lost_to_filtering}

qc.spec <- list(
  nCount_RNA   = c(lower = RNA.count.floor,     upper = RNA.count.ceiling),
  nFeature_RNA = c(lower = feature.count.floor, upper = feature.count.ceiling),
  percent.mt   = c(upper = mito.ceiling),
  percent.hb   = c(upper = rbc.ceiling),
  percent.ribo = c(lower = ribo.floor))



# render_plain_table(qc_attrition(md.pre, qc.spec), caption = "Cells excluded by each QC criterion")

qc.attrition.table <- qc_attrition(md.pre, qc.spec)
render_plain_table(qc.attrition.table,
                   caption = "Cells excluded by each QC criterion")

to.save.for.report.p2 <- c("qc.attrition.table", "qc.spec", "combo.table", "md.pre")
save(list = to.save.for.report.p2, file = part2.tables.path)

# client-facing copy, alongside filtering_stats_table.csv
write.csv(qc.attrition.table,
          file.path(r.obj.loc, "qc_attrition_table.csv"),
          row.names = FALSE, quote = FALSE)
```


```{r filtered_metric_tables}

md <- seurat5.obj[[]]
sample_col <- if ("sample_id" %in% colnames(md)) "sample_id" else {
  # fallback: infer sample from cell names like "SAMPLE_barcode"
  seurat5.obj$orig.ident <- sub("_.*$", "", colnames(seurat5.obj))
  "orig.ident"
}

# --- build the same tables you had before, now from the merged & filtered object ---
nCount.data.filtered      <- make_metric_table(seurat5.obj, sample_col, "nCount_RNA")
nFeature.data.filtered    <- make_metric_table(seurat5.obj, sample_col, "nFeature_RNA")
percent.mt.data.filtered  <- make_metric_table(seurat5.obj, sample_col, "percent.mt")
percent.hb.data.filtered  <- make_metric_table(seurat5.obj, sample_col, "percent.hb")
percent.y.data.filtered   <- make_metric_table(seurat5.obj, sample_col, "percent.y")
percent.ribo.data.filtered <- make_metric_table(seurat5.obj, sample_col, "percent.ribo")

# For custom sorting order in violin plots:
# sample_order <- c("sample_1", "sample_2",
#                   "smaple_3", "sample_4")
# nCount.data.filtered <- make_metric_table(seurat5.obj, sample_col, "nCount_RNA", sample_levels = sample_order)


```

# QC filtering, before and after

Rows are metrics; the left panel is raw with the filtered range shaded red, the
right panel is what survives. Both panels in a row share a y-axis, so the
comparison is direct. A cutoff that removes no cells is drawn without shading.

## Library size and complexity

```{r qc_ba_counts}
qc.counts <- rbind(
  qc_long(list(nCount_RNA = nCount_data,
               nFeature_RNA = nFeature_data),          stage = "raw"),
  qc_long(list(nCount_RNA = nCount.data.filtered,
               nFeature_RNA = nFeature.data.filtered), stage = "filtered"))

thr.counts <- qc_thresholds(
  list(nCount_RNA   = c(lower = RNA.count.floor,     upper = RNA.count.ceiling),
       nFeature_RNA = c(lower = feature.count.floor, upper = feature.count.ceiling)),
  metric.levels = levels(qc.counts$metric))

p.qc.counts <- qc_facet_plot(qc.counts, thresholds = thr.counts, log.y = TRUE)
p.qc.counts

save_png_plot(p.qc.counts, 
              plot.path(ggplot.directory, "qc_counts_before_after.png", part2.suffix),
              width = 6.5, height = 5)
```

## Contamination metrics

```{r qc_ba_percents}
qc.pct <- rbind(
  qc_long(list(percent.mt = percent_mt_data,
               percent.hb = percent_hb_data,
               percent.ribo = percent_ribo_data),           stage = "raw"),
  qc_long(list(percent.mt = percent.mt.data.filtered,
               percent.hb = percent.hb.data.filtered,
               percent.ribo = percent.ribo.data.filtered),  stage = "filtered"))

thr.pct <- qc_thresholds(
  list(percent.mt   = c(upper = mito.ceiling),
       percent.hb   = c(upper = rbc.ceiling),
       percent.ribo = c(lower = ribo.floor)),
  metric.levels = levels(qc.pct$metric))

p.qc.pct <- qc_facet_plot(qc.pct, thresholds = thr.pct)
p.qc.pct
save_png_plot(p.qc.pct,
              plot_path(ggplot.directory, "qc_percent_before_after.png", part2.suffix),
              width = 9, height = 11)
```

# Analysis without integration


```{r indiv_samples_sctransform_doublet_removal, eval=run.doubletFinder}
t0 <- Sys.time()

sample.list <- SplitObject(seurat5.obj, split.by = "sample_id")

sample.list <- lapply(sample.list, function(obj) {
  # SCTransform + dim reduction
  obj <- SCTransform(obj, verbose = FALSE)
  obj <- RunPCA(obj, verbose = FALSE)
  obj <- RunUMAP(obj, dims = 1:30, verbose = FALSE)
  obj <- FindNeighbors(obj, dims = 1:30, verbose = FALSE)
  obj <- FindClusters(obj, verbose = FALSE)

  # pK identification
  sweep.res   <- paramSweep(obj, PCs = 1:10, sct = TRUE)
  sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
  bcmvn       <- find.pK(sweep.stats)
  pk.val      <- as.numeric(as.character(bcmvn$pK[which.max(bcmvn$BCmetric)]))

  # Homotypic adjustment using fresh clusters
  annotations    <- obj@meta.data$seurat_clusters
  homotypic.prop <- modelHomotypic(annotations)
  nExp_poi       <- round(0.075 * nrow(obj@meta.data))
  nExp_poi.adj   <- round(nExp_poi * (1 - homotypic.prop))

  sample.name <- unique(obj$sample_id)

  cat(sample.name, " nCells:", ncol(obj), "| pK:", pk.val,
      "| nExp:", nExp_poi, "| nExp.adj:", nExp_poi.adj, "\n")

  # DoubletFinder - run 1 (unadjusted)
  obj      <- doubletFinder(obj, PCs = 1:10, pN = 0.25, pK = pk.val,
                             nExp = nExp_poi, reuse.pANN = NULL, sct = TRUE)
  pann.col <- grep("pANN", names(obj[[]]), value = TRUE)

  # DoubletFinder - run 2 (homotypic adjusted)
  obj      <- doubletFinder(obj, PCs = 1:10, pN = 0.25, pK = pk.val,
                             nExp = nExp_poi.adj, reuse.pANN = pann.col, sct = TRUE)
  df.col   <- tail(grep("DF.classifications", names(obj[[]]), value = TRUE), 1)

  cat("Doublets removed:", table(obj[[df.col]])[["Doublet"]],
      "| Singlets retained:", table(obj[[df.col]])[["Singlet"]], "\n")

  # Remove doublets
  singlet.cells <- rownames(obj@meta.data)[obj@meta.data[[df.col]] == "Singlet"]
  subset(obj, cells = singlet.cells)
})

# Re-merge cleaned samples
seurat5.obj <- merge(sample.list[[1]], y = sample.list[-1])

time.len <- difftime(Sys.time(), t0, units = "mins")
print(paste0("SCTransform + DoubletFinder took ", round(time.len, 2), " minutes to run."))
saveRDS(seurat5.obj, merged.seurat.obj.path)
print(paste0("Saved merged object to ", merged.seurat.obj.path))


```

```{r doubletFinderMsg, eval=run.doubletFinder, results='asis'}
cat("## Doublet Detection\n\nDoubletFinder was run on each sample individually prior to merging.\n")
```

SCTransform, PCA and UMAP generation

```{r process_merged_object}

# seurat5.obj <- readRDS(merged.seurat.obj.path)

t0 <- Sys.time()

seurat5.obj <- SCTransform(seurat5.obj)

time.len <- difftime(Sys.time(), t0, units = "mins")
print(paste0("SCTransform took ", round(time.len, 2), " minutes to run."))

seurat5.obj <- RunPCA(seurat5.obj)
seurat5.obj <- RunUMAP(seurat5.obj, dims = 1:30)

saveRDS(seurat5.obj, merged.seurat.obj.path)
print(paste0("Saved merged object to ", merged.seurat.obj.path))
```

## Umap of merged, non-integrated object

```{r merged_obj_umap}

# seurat5.obj <- readRDS(merged.seurat.obj.path)

merged.dimplot <- DimPlot(seurat5.obj, 
                          reduction = "umap", 
                          group.by = "sample_id",
                          shuffle = TRUE,
                          raster=FALSE)


save_png_plot(merged.dimplot, 
              plot.path(ggplot.directory, "umap_merged_seruat_obj.png", part2.suffix)
merged.dimplot
```



# Integration

```{r integrate_seurat_obj}
integrated.reduction.label <- "integrated.rpca"

options(future.globals.maxSize = +Inf)
# seurat5.obj <- readRDS(merged.seurat.obj.path)

# --- START cell num check ---
# Calculate the minimum number of cells in any layer/sample
min.cells <- min(table(seurat5.obj$sample_id))

# Set k.weight to the minimum of 100 (default) or 70% of the smallest sample size
# assuming a worst-case where only 70% of the cells end up acting as anchors
if(min.cells < 150) {
  k.weight.val <- floor(min.cells * 0.7)
  message(paste("Warning: Smallest sample has", min.cells, "cells."))
  message(paste("Setting k.weight conservatively to", k.weight.val, "to avoid anchor error."))
} else {
  k.weight.val <- 100
}

message(paste("Smallest sample size is", min.cells, " - setting k.weight to ", k.weight.val))


t0 <- Sys.time()
seurat5.obj.integrated <- IntegrateLayers(
  object = seurat5.obj,
  method = RPCAIntegration,
  normalization.method = "SCT",
  orig.reduction = "pca",
  new.reduction = integrated.reduction.label,
  k.weight = k.weight.val, # <--- Pass the dynamic value here
  verbose = FALSE
)

time.len <- difftime(Sys.time(), t0, units = "mins")

print(paste0("IntegrateLayers took ", round(time.len, 2), " minutes to run."))
# re-join layers after integration
seurat5.obj.integrated[["RNA"]] <- JoinLayers(seurat5.obj.integrated[["RNA"]])

# Build the SNN graph on the integrated embedding under a STABLE name
# ("int_snn"), so part 2b can cluster straight off the saved object without
# recomputing neighbours.
seurat5.obj.integrated <- FindNeighbors(seurat5.obj.integrated,
                                        reduction = integrated.reduction.label, 
					dims = 1:30,
					graph.name = c("int_nn", "int_snn"))

# # --- post-integration UMAP (note the reduction name) ---
seurat5.obj.integrated <- RunUMAP(
  seurat5.obj.integrated, dims = 1:30, reduction = integrated.reduction.label,
  reduction.name = "umap.postint"
)


# Stamp the upstream parameters into the object so part 2b can detect a stale
# handoff (i.e. an integrated object built under different QC / integration /
# doublet settings than the current config).
seurat5.obj.integrated@misc$upstream_params <- upstream_params(
  p2               = p2,
  run.doubletFinder = run.doubletFinder,
  organism         = organism,
  part1.suffix     = part1.suffix,
  part2.suffix     = part2.suffix
)

saveRDS(seurat5.obj.integrated, integrated.seurat.obj.path, compress = FALSE)
print(paste0("Saved integrated object (with int_snn graph) to ", integrated.seurat.obj.path))
```

# Clustering resolution recommendation

Part 2a does **not** commit to a final clustering. It scores a grid of
resolutions for *subsample stability* — how reproducibly each resolution
recovers the same clusters when a fraction of cells is resampled — and writes a
**proposed** resolution to the config. Part 2b reads that value (or your manual
override) and does the actual clustering + marker discovery.

Stability is chosen over silhouette because silhouette systematically rewards
coarse partitions; stability instead answers the question we actually care about
— the finest resolution the data still supports reproducibly.

```{r stability_sweep}
# Resolution grid (config-driven; shared with the old silhouette sweep).
res.min  <- if (!is.null(p2$cluster_res_min))  as.numeric(p2$cluster_res_min)  else 0.1
res.max  <- if (!is.null(p2$cluster_res_max))  as.numeric(p2$cluster_res_max)  else 1.0
res.step <- if (!is.null(p2$cluster_res_step)) as.numeric(p2$cluster_res_step) else 0.1
resolutions <- round(seq(res.min, res.max, res.step), 3)

# Stability knobs (config-driven, with sensible defaults).
n.sub  <- if (!is.null(p2$stability_reps))           as.integer(p2$stability_reps)           else 5L
frac   <- if (!is.null(p2$stability_subsample_frac)) as.numeric(p2$stability_subsample_frac) else 0.8
thresh <- if (!is.null(p2$stability_threshold))      as.numeric(p2$stability_threshold)      else 0.75

t0 <- Sys.time()
stability    <- select_resolution_by_stability( 
  obj = seurat5.obj.integrated,
  resolutions = resolutions, 
  graph.name = "int_snn",
  reduction = integrated.reduction.label, 
  dims = 1:30, 
  n_subsample = n.sub,
  subsample_frac = frac, 
  seed = 1)   # no stability_threshold

stability.df <- stability$summary
proposed.res <- stability$chosen_resolution

cat("Data-derived threshold:", round(stability$threshold, 3),
    "| proposed resolution:", proposed.res, "\n")

message("Stability sweep took ", round(difftime(Sys.time(), t0, units = "mins"), 2), " min")

stability.df <- stability$summary
proposed.res <- stability$chosen_resolution

print(stability.df)
cat("\nProposed resolution (finest that stays >= ", thresh, " stability): ",
    proposed.res, "\n", sep = "")
```

```{r stability_plot}
# chooseR-style robustness plot: per-resolution MEDIAN of the per-cluster
# stability distribution with a bootstrap 95% CI (crossbars). The threshold is
# DATA-DERIVED (highest lower-CI bound across non-trivial resolutions, blue
# line); the chosen resolution is the finest one whose median clears it (red).
thr        <- stability$threshold
lvls       <- levels(factor(stability.df$resolution))
chosen.pos <- match(as.character(proposed.res), lvls)

p.stab <- ggplot(stability.df, aes(x = factor(resolution), y = median)) +
  geom_crossbar(aes(ymin = low, ymax = high), fill = "grey85", linewidth = 0.25) +
  geom_hline(yintercept = thr, colour = "steelblue") +
  geom_vline(xintercept = chosen.pos, colour = "firebrick") +
  geom_text(aes(label = n_clusters), vjust = -0.8, size = 3) +
  labs(
    title    = "Cluster robustness vs. resolution (data-derived threshold)",
    subtitle = paste0("numbers = cluster count | blue = threshold (",
                      round(thr, 3), ") | red = chosen (", proposed.res, ")"),
    x = "resolution",
    y = "median per-cluster stability (best-match Jaccard), 95% CI"
  ) +
  theme_minimal()

save_png_plot(p.stab,
              plot_path(ggplot.directory, "stability_vs_resolution.png", part2.suffix),
              width = 8, height = 6)
p.stab
```

```{r clustree_plot, eval = requireNamespace("clustree", quietly = TRUE)}
# Optional override aid: how clusters split/merge across the grid. Skipped
# automatically if the clustree package isn't installed. NOTE: this recomputes
# FindClusters across the grid to populate the int_snn_res.* columns; it could
# be merged with the stability sweep later to avoid the double pass.
for (res in resolutions) {
  seurat5.obj.integrated <- FindClusters(seurat5.obj.integrated,
                                         graph.name = "int_snn",
                                         resolution = res, verbose = FALSE)
}
clustree.plot <- clustree::clustree(seurat5.obj.integrated, prefix = "int_snn_res.")
save_png_plot(clustree.plot,
              plot_path(ggplot.directory, "clustree.png", part2.suffix),
              width = 9, height = 11)
clustree.plot
```

```{r write_proposed_resolution}
# Write the PROPOSED resolution back to config as the value users can override.
# Part 2b derives the label and writes the authoritative num_clusters after it
# actually clusters; only the numeric value is written here.
cfg$analysis$part2$clustering_resolution_value <- proposed.res
yaml::write_yaml(cfg, params$config.args)
cat("Wrote proposed clustering_resolution_value =", proposed.res, "to", params$config.args, "\n")
cat("To override: edit clustering_resolution_value in config.yaml, then run part 2b.\n")
```

## Session Information
```{r session_info_2a}
sessionInfo()
```
