---
title: "scRNA-seq Summary Report"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    includes:
      in_header: common-pdf.tex
params:
  config.args: "./config.yaml"
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning=FALSE,error=FALSE, message=FALSE, fig.align='center', out.width='\\linewidth'
)
# source helpers that sit beside this Rmd
source("functions_for_report.R")
```



```{r import_libraries}
# must have Seurat. amd bookdown, yaml, ggplot2, and tinytex
# suppressMessages(require(Seurat))
# suppressMessages(require(ggplot2))
if (!suppressMessages(require(yaml))) stop("yaml is required")
if (!suppressMessages(require(scales))) stop("scales is required")
if (!suppressMessages(require(kableExtra))) stop("kableExtra is required")
if (!suppressMessages(require(dplyr))) stop("dplyr is required")
if (!suppressMessages(require(ggplot2))) stop("ggplot2 is required")

# Only loading this so that it shows up in the sessionInfo part of the report
if (!suppressMessages(require(Seurat))) stop("Seurat is required") 
```


```{r user_vars}

# get the datetime and set it to variable for inline printing
todays.date <- Sys.Date()
formatted.date <- format(todays.date, format="%B %d %Y")

# Import the config file
config.args <- read_yaml(params$config.args)

run_metrics_tables=TRUE
metrics.table.string <- config.args$analysis$part1$generate_metrics_tables

if (isFALSE(as.logical(metrics.table.string %||% TRUE))){
  run_metrics_tables=FALSE
}

cfg <- config.args
cfg.analysis <- cfg$analysis

p1  <- cfg.analysis$part1
p2 <- cfg.analysis$part2

################################################################################
# set project variables
################################################################################
analyst.name <- config.args$project$`analyst-name`
formatted.samples <- config.args$project$samples # Sanitized below
experiment.name = config.args$project$`project-name` 
project.description <- config.args$project$`project-description`
dataset.path <- config.args$data$parent_directory_path
organism = config.args$project$organism

################################################################################
# Input paths. Part 3 reads from all three stages; see stage_dirs().
################################################################################
dirs <- stage_dirs(config.args, with.part2b = TRUE)
if (!dir.exists(dirs$part2b))
  stop("Part 2b output not found at ", dirs$part2b,
       " - run part 2b (at resolution ", dirs$resolution, ") first.")

QC.tables.path    <- file.path(dirs$part1,  p1$part1_report_tables_filename)
part2.tables.path <- file.path(dirs$part2a, p2$part2_report_tables_filename)

ggplot.dir.p1   <- file.path(dirs$part1,  p1$ggplot_dir)   # no consumer yet
ggplot.dir.p2a  <- file.path(dirs$part2a, p1$ggplot_dir)
ggplot.dir.p2b  <- file.path(dirs$part2b, p1$ggplot_dir)
marker.gene.dir <- file.path(dirs$part2b, p2$marker_gene_dir)

ids <- sanitize_sample_names(formatted.samples)
```


## Report and variable details

This summary report was generated on `r formatted.date`

* organism: `r organism`

## Project Name

`r experiment.name`

## Project Description:

`r project.description`

## Sample Names

`r paste(paste0("- ", formatted.samples), collapse = "\n")`


# Part 1: QC

## Sequencing Metrics

For more insight on what the different metrics presented here actually mean, see this article on the 10x website: https://www.10xgenomics.com/analysis-guides/quality-assessment-using-the-cell-ranger-web-summary

<br/>
<br/>

```{r prep_qc_table, eval=run_metrics_tables, message=FALSE, warning=FALSE}
# --- Import the metrics summaries (quiet) ---
qc.table <- make_qc_table(ids, dataset.path)
render_formatted_table(qc.table)
```

\blandscape

## scRNA-seq QC

Distributions before and after filtering. Left column raw, red shading marks
the excluded range; right column post-filter. Cell counts per sample are in the filtering table below.



```{r qc_figures, out.width='100%'}
knitr::include_graphics(vapply(
  c("qc_counts_before_after.png", "qc_percent_before_after.png"),
  function(x) plot_path(ggplot.dir.p2a, x), character(1), USE.NAMES = FALSE))
```

\elandscape

\blandscape

## Filtering results


Here is a table showing the number of cells before applying the filters described above,
and then the number of cells remaining after filtering. 

```{r filtering_res_table, results='asis'}

combo.table <-  read.csv(file.path(dirs$part2a, "filtering_stats_table.csv"))

colnames(combo.table)[1] <- "Metric"

combo.table %>%
  kable(format = "latex",
        booktabs = TRUE,
        caption = "Cell counts before and after filtering",
        row.names = FALSE) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size = 8
  ) %>%
  row_spec(0, angle = 45, font_size = 7) %>%
  column_spec(1, bold = TRUE)

```


Table showing the number of cells lost to each qc cutoff:

```{r qc_attrition_table}
load(part2.tables.path)

render_plain_table(qc.attrition.table,
                   caption = "Cells excluded by each QC criterion")
```

\elandscape

# Part 2: Integration, clustering, and differential expression

## Pre-integration UMAP

```{r pre-int_umap}
merged.umap.path <- plot_path(ggplot.dir.p2a, "umap_merged_seruat_obj.png")

knitr::include_graphics(merged.umap.path)
```

### Post-integration UMAP

We integrate all of the samples together into a single object for comparison
of cell types across samples. Go [here](https://satijalab.org/seurat/articles/integration_introduction.html) to read more about single-cell integration. 

## Post-integration UMAP by sample

```{r post-int_umap}
integrated.umap.path <- plot_path(ggplot.dir.p2a, "integrated_umap_by_sample.png")

knitr::include_graphics(integrated.umap.path)
```


### Clustering

To group cells into clusters, we scan a range of clustering resolutions (a resolution setting simply controls how finely the cells get divided: lower values yield fewer, broader clusters and higher values yield more, smaller ones). To choose among them, we measure cluster stability: at each resolution we repeatedly take a random 80% of the cells, re-cluster them, and check how consistently the same clusters reappear. Clusters that keep reforming under this resampling are treated as real structure, while ones that dissolve are a sign the data is being split too finely. 

We then automatically select the finest resolution whose clusters remain reliably reproducible, which gives us as much biological detail as the data can actually support without over-splitting. This is an objective starting point rather than a final answer: the "best" resolution is flexible, since neighboring values often give very similar results, and interpretation still depends on marker genes and biological context. In other words, this clustering is an educated guess that can be adjusted (clusters can be merged or split, or a different resolution chosen to generate more or fewer clusters, though the overall shape of the UMAPs won't change).


The plot below shows this stability at each resolution (higher is more stable), with the number of clusters labeled above each point. The pipeline selected the resolution marked in red: the finest one whose stability still clears the data-derived threshold shown in blue.


## Cluster stability by resolution

```{r cluster_stability}
clust.stability.res.path <- plot_path(ggplot.dir.p2a, "stability_vs_resolution.png")

knitr::include_graphics(clust.stability.res.path)
```


## Post-integration UMAP by cluster

```{r clusters_by_res}
cluster.umap.path <- plot_path(ggplot.dir.p2b, "integrated_umap_by_cluster.png")

knitr::include_graphics(cluster.umap.path)
```

\pagebreak

## Table of cells per cluster from each sample

```{r cluster_sample_table, results='asis'}

cluster.sample.df <- read.csv(file.path(dirs$part2b, "cells_perCluster_perSample.csv"),
                              row.names=1)
cluster.sample.df %>%
  kable(format = "latex",
        booktabs = TRUE,
        caption = "Cell counts per cluster per sample",
        longtable = FALSE,
        row.names = FALSE) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size = 8
  ) %>%
  row_spec(0, angle = 45, font_size = 7) %>%
  row_spec(nrow(cluster.sample.df), bold = TRUE) %>%  # bold Total row
  column_spec(1, bold = TRUE) %>%                      # bold Cluster column
  column_spec(ncol(cluster.sample.df), bold = TRUE)    # bold Total column

# cluster.sample.df
```


\pagebreak

## Marker genes

For each cluster, which genes are differentially expressed in that 
cluster compared to all other clusters combined?

```{r gene_feature, fig.width = 25, fig.height = 20}
assay.used <- "SCT"

# Cluster IDs come from the marker table 2b wrote, not 0:(num_clusters - 1):
# a cluster with no positive markers gets no figures (only.pos = TRUE), and IDs
# aren't guaranteed contiguous if clusters were filtered by hand.
marker.csv <- file.path(marker.gene.dir, "cluster_markers.csv")
if (!file.exists(marker.csv))
  stop("Marker table not found: ", marker.csv, " — run part 2b first.")
clusters.present <- sort(unique(read.csv(marker.csv)$cluster))

# Mirror the plot types 2b was configured to draw, so part 3 never asks for a
# panel that was never generated.
marker.plot.types <- unlist(p2$marker_plot_types) %||% c("feat", "dot")

# One path per (cluster, kind), interleaved so each cluster's panels stay together.
figure.list <- unlist(lapply(clusters.present, function(cluster.num) {
  vapply(marker.plot.types,
         function(kind) marker_plot_path(ggplot.dir.p2b, cluster.num, kind,
                                         assay.used),
         character(1))
}), use.names = FALSE)

absent <- figure.list[!file.exists(figure.list)]
if (length(absent))
  warning("Marker plots missing, omitted from the report: ",
          paste(basename(absent), collapse = ", "))
figure.list <- figure.list[file.exists(figure.list)]

if (length(figure.list) == 0) {
  cat("No marker gene plots were found in", ggplot.dir.p2b, "\n")
} else {
  knitr::include_graphics(figure.list)
}
```


## Analysis parameters

```{r run_parameters}
run.params <- read.csv(file.path(dirs$part2b, "run_parameters.csv"),
                       stringsAsFactors = FALSE)
render_plain_table(run.params, caption = "Analysis parameters for this run")
```


## Session Information
```{r sessioinfo}
sessionInfo()
```


